---
title: "Assignment 5"
author: "Denis Cruz"
date: "2026-04-06"
format:
  html:
    toc: true
execute:
  warning: false
  message: false
---

## Introduction

In this assignment, we analyz socio ecological monitoring data from the Sacramento San Joaquin Delta. The goal was to visualize visitor activity across restoration locations and map the spatial distribution, these visualizations help us visualize human activity and geographic patterns across the Delta ecosystem

```{r}
# loading in necessary packages
library(readr)
library(dplyr)
library(tidyr)
library(forcats) # makes working with factors easier
library(ggplot2)
library(leaflet) # interactive maps
library(DT) # interactive tables
library(scales) # scale functions for visualization
library(janitor) # expedite cleaning and exploring data
library(viridis) # colorblind friendly color palette

```

```{r}
# read in data from the data directory after manually downloading data 
delta_visits_raw <- read_csv(here::here("assignments/data/raw/Socioecological_monitoring_data.csv"))
```

```{r}
#looking into raw data 
## Check out column names

colnames(delta_visits_raw)

## Peak at each column and class
glimpse(delta_visits_raw)

## From when to when
range(delta_visits_raw$Date)

## Which time of day?
unique(delta_visits_raw$Time_of_Day)
```

```{r}
# generalizing names 
delta_visits <- delta_visits_raw %>% 
    janitor::clean_names()
colnames(delta_visits)

```

```{r}
visits_long <- delta_visits %>% 
    pivot_longer(cols = c(sm_boat, med_boat, lrg_boat, bank_angler, scientist, cars),
                 names_to = "visitor_type",
                 values_to = "quantity") %>%
    rename(restore_loc = eco_restore_approximate_location) %>% 
    select(-notes)

## Checking the outcome
head(visits_long)

daily_visits_loc <- visits_long %>%
    group_by(restore_loc, date, visitor_type) %>% 
    summarise(daily_visits = sum(quantity))
    
head(daily_visits_loc)
```

```{r}
my_theme <- theme_bw(base_size = 16) +
    theme(legend.position = "bottom",
          axis.ticks.y = element_blank())

```

```{r}
daily_visits_totals <- daily_visits_loc %>% 
    group_by(restore_loc) %>%
    mutate(n = sum(daily_visits)) %>% 
    ungroup() %>%
    mutate(restore_loc = fct_reorder(restore_loc, n))

head(daily_visits_totals)
```

```{r}
levels(daily_visits_totals$restore_loc) ### not alphabetical any more!
```

```{r}
## Graph with comments 

daily_visits_totals <- daily_visits_loc %>% 
    group_by(restore_loc) %>% # this groups by resoration locations
    mutate(n = sum(daily_visits)) %>% #adds total visits per location and adds a column
    ungroup() %>% #no grouping ? 
    mutate(restore_loc = fct_reorder(restore_loc, desc(n))) #re order location by visit most to least

ggplot(data = daily_visits_totals,
       aes(x = daily_visits, y = restore_loc,
           fill = visitor_type)) +
    geom_col() +
    labs(x = "Number of Visits", # X axis
         y = "Restoration Location", #y axis
         fill = "Type of Visitor", #color by visit type
         title = "Total Number of Visits to Delta Restoration Areas by visitor type", #title will appear at the top
         subtitle = "Sum of all visits during study period") + #subtitle
    scale_x_continuous(breaks = seq(0,120, 20), expand = c(0,0)) +
    my_theme #x axis numner 0 to 120 with 20 intervals #expand removes extra space in graph #my_theme adds the custom theme we made before
```

```{r}
# DT interactive tables 
locations <- visits_long %>%
    distinct(restore_loc, .keep_all = TRUE) %>%
    select(restore_loc, latitude, longitude)

head(locations)
```

```{r}
datatable(locations)
```

```{r}
## interactive map with comments

leaflet(locations) %>% #open map with thelocation data
    addProviderTiles(providers$Esri.WorldImagery) %>% #adds satellite imaging
    addProviderTiles(providers$CartoDB.PositronOnlyLabels) %>% #adds lables location names
    addCircleMarkers(
        lng = ~ longitude, #longitude for marker
        lat = ~ latitude, #latitude for marker
        popup = ~ restore_loc, # shows location name when clicked
        radius = 5, #marker size
        # set fill properties
        fillColor = "salmon", #marker color
        fillOpacity = 1, #opacity 
        # set stroke properties
        stroke = TRUE, #allows border around marker
        weight = 0.5, #border thickeness
        color = "white", #white border
        opacity = 1) #opacity 
```

## Discussion

The bar plot shows how the total number of visitors varies across locations. We can see that some sites receive higher levels of activity than others, indicating that human use is not distributed evenly across the Delta. The stacked bars are helful because they show that different visitor types contribute differently at each location. For example, some sites seem to have more boat activity, while others have more cars or bank anglers, suggesting differences in how each site is accessed and used, which may be better explained looking at are map. \
\
The map provides important context for the patterns we see in the bar plot by showing where each restoration location is in the Delta. The sites are spread across a wide geographic area, which helps see reason of visitation type. for example locations that are more accessible or closer to people may receive more visitors, while more remote sites may see less activity. Which is where i'd be going. \
\
Together, the plot and the map complement each other by combining quantitative patterns with location context. The plot summarizes how visitation differs across sites and visitor types and the map helps explain these differences by showing where the sites are located.
